\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-09-06 NAR_DRAFT.tex
paragraph
figure changes

\begin{figure}
  \begin{center}
  \includegraphics{figures/signature_figure.pdf}
  \end{center}
  \caption{Mutation profiles. 
  \textbf{(a)} The observed mutation profile of a breast cancer sample from \cite{Shi2018}.
  \textbf{(b)} The observed mutation profile of the AML 31 primary tumor from \cite{Griffith2015}.
  \textbf{(c)} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
  \textbf{(d)} Equal proportions of signatures 1, 4, \& 5.
  \textbf{(e)} Equal proportions of signatures 1, 3, \& 5.
  KL divergence between the estimated profile ($\boldsymbol{\pi}$) and the simulated mutation profile for \textbf{(f)} Whole exomes and \textbf{(g)} whole genomes.
  The estimated mutation profile converges quickly to the target as the number of mutations evaluated rises.
  }
  \label{NAR-sigfig}
  \end{figure}

@@

\begin{equation}
  \label{eqn:computed_posterior}
  $$
  \frac{\mathrm{P}(m,\nu = \hat{f})}{1 - \mathrm{P}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} + \mathrm{logit}_{10}(\mathrm{p}(\mathrm{S}_{m,c}))},
  $$
\end{equation}
where $\mathrm{p}(\mathrm{S}_{m,c})$ is the prior probability of a substitution of type $\mathrm{S}_{m,c}$, as described in equation \ref{eqn:detailed_prior} and specified in equations \ref{eqn:post_pred} - \ref{eqn:mut_rate}.
When comparing our posterior odds to those of MuTect in this paper, we use the assumption put forward by the MuTect authors in \citet{Cibulskis2013}.
They suggest a threshold for the posterior odds of 2, and a prior per-base probability of mutation of $3\cdot10^{-6}$, leading to a \texttt{TLOD} threshold of 6.3.
We compute the posterior odds for Mutect as
\begin{equation}
  \label{eqn:mutect_posterior}
  $$
  \frac{\mathrm{P}(m,\nu = \hat{f})}{1 - \mathrm{P}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} - 6},
  $$
\end{equation}
since $\mathrm{logit}_{10}(3\cdot10^{-6}) \approx -6$, and $10^{.3} \approx 2$.

@@

@@

We have chosen to implement \batcave as a post-calling algorithm for exclusive use with MuTect because MuTect presents a simple model and a well-tested set of heuristic filters.
However, the model is very general, and in principle there is no barrier to using the method with other variant callers.
For instance Strelka2, another widely used variant caller, computes a joint posterior probability over tumor and normal genotypes where the prior probability that the tumor carries a somatic mutation at a given site is set to a constant \cite{Kim2018}.
Modifying the algorithm to work with the Strelka statistical model would involve replacing the constant prior with a tumor and site-specific prior for each potential variant, exactly as we have done with MuTect.
The R implementation \texttt{batcaver} adds 1 second per 22,000 variants evaluated to a standard GATK best practices variant calling pipeline. 
The majority of the computational overhead in \batcave is associated with extracting the tri-nucleotide context for each potential variant site from the reference genome.
Since most variant callers are already walking the reference genome as part of the calling process, extracting the tri-nucleotide context simultaneously would dramatically reduce the computational overhead associated with implementing \batcave.

>>>>

\begin{figure}
  \begin{center}
  \includegraphics{figures/signatures_only.pdf}
  \end{center}
  \caption{Mutation profiles. 
  \textbf{(a)} The observed mutation profile of a breast cancer sample from \cite{Shi2018}.
  \textbf{(b)} The observed mutation profile of the AML 31 primary tumor from \cite{Griffith2015}.
  \textbf{(c)} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
  \textbf{(d)} Equal proportions of signatures 1, 4, \& 5.
  \textbf{(e)} Equal proportions of signatures 1, 3, \& 5.
  }
  \label{NAR-sigfig}
  \end{figure}

@@

\begin{equation}
  \label{eqn:computed_posterior}
  $$
  \frac{\mathrm{P}(m,\nu = \hat{f})}{1 - \mathrm{P}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} + \mathrm{logit}_{10}(\mathrm{p}(\mathrm{S}_{m,c}))},
  $$
\end{equation}
where $\mathrm{p}(\mathrm{S}_{m,c})$ is the prior probability of a substitution of type $\mathrm{S}_{m,c}$, as described in equation \ref{eqn:detailed_prior} and specified in equations \ref{eqn:post_pred} - \ref{eqn:mut_rate}.
When comparing our posterior odds to those of MuTect in this paper, we use \citet{Cibulskis2013} in assuming a prior per-base probability of mutation of $3\cdot10^{-6}$.
We compute the posterior odds for Mutect as
\begin{equation}
  \label{eqn:mutect_posterior}
  $$
  \frac{\mathrm{P}(m,\nu = \hat{f})}{1 - \mathrm{P}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} - 6},
  $$
\end{equation}
since $\mathrm{logit}_{10}(3\cdot10^{-6}) \approx -6$.

@@

\begin{figure}
  \begin{center}
  \includegraphics{figures/kl_only.pdf}
  \end{center}
  \caption{KL divergence between the estimated profile ($\boldsymbol{\pi}$) and the simulated mutation profile for \textbf{(f)} Whole exomes and \textbf{(g)} whole genomes.
  The estimated mutation profile converges quickly to the target as the number of mutations evaluated rises.}
\label{NAR-ppv_fig}
\end{figure}

@@

We have chosen to implement \batcave as a post-calling algorithm for exclusive use with MuTect because MuTect presents a simple model and a well-tested set of heuristic filters.
However, the model is very general, and in principle there is no barrier to using the method with other variant callers.
For instance Strelka2, another widely used variant caller, computes a joint posterior probability over tumor and normal genotypes where the prior probability that the tumor carries a somatic mutation at a given site is set to a constant \cite{Kim2018}.
Modifying the algorithm to work with the Strelka statistical model would involve replacing the constant prior with a tumor and site-specific prior for each potential variant, exactly as we have done with MuTect.

The R implementation \texttt{batcaver} adds 1 second per 22,000 variants evaluated to a standard GATK best practices variant calling pipeline. 
The majority of the computational overhead in \batcave is associated with extracting the tri-nucleotide context for each potential variant site from the reference genome.
Since most variant callers are already walking the reference genome as part of the calling process, extracting the tri-nucleotide context simultaneously would dramatically reduce the computational overhead associated with implementing \batcave.
